Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-08-10 Paper/Paper.tex
changed abstract and added limitations of capacitive sensors, deanonymized it

% the name(s) of the author(s) follow(s) next
% \author{Tobias Grosse-Puppendahl\inst{1} \and Eugen Berlin\inst{2} \and Marko Borazio\inst{2} }
\author{Author 1 \and Author 2 \and Author 3 }

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
% \institute{
% 	Fraunhofer IGD, Fraunhoferstr. 5, 64283 Darmstadt, Germany\\
% 	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
% 	\and
% 	Technische Universit\"at Darmstadt, Hochschulstr. 10, 64289 Darmstadt, Germany\\
% 	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
% }

\institute{
	Anonymized for review \\
	\email{anonymous@insitution.edu}
	\and
	Anonymized for review \\
	\email{anonymous@insitution.edu}
}

% \toctitle{Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing}
% \tocauthor{Eugen Berlin, Marko Borazio, Tobias Grosse-Puppendahl}

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the user's environment to boost activity recognition. This work presents a sensor prototype consisting of an accelerometer and a capacitive proximity sensor that senses the user's activities based on the combined sensor values. We show that our proposed approach significantly improves the recognition rate for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
We obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{Wimmer2007} who presented a toolkit for capacitive proximity sensing.

In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{When the participants entered the apartment, the wristband approached the door knob twice, at the time of opening and closing the door. This fact can be observed in the capacitive proximity data (upper plot) at the beginning and at the end of the activity, whereas the acceleration has no characteristic information (bottom plot).}
	\label{fig:opendoor}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

@@

Anonymized for review.

% We would like to thank our evaluation participants from TU Darmstadt and Fraunhofer IGD.

>>>>

% the name(s) of the author(s) follow(s) next
\author{Tobias Grosse-Puppendahl\inst{1} \and Eugen Berlin\inst{2} \and Marko Borazio\inst{2} }
%\author{Author 1 \and Author 2 \and Author 3 }

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{
 	Fraunhofer IGD, Fraunhoferstr. 5, 64283 Darmstadt, Germany\\
 	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
 	\and
 	Technische Universit\"at Darmstadt, Hochschulstr. 10, 64289 Darmstadt, Germany\\
 	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
}

%\institute{
	%Anonymized for review \\
	%\email{anonymous@insitution.edu}
	%\and
	%Anonymized for review \\
	%\email{anonymous@insitution.edu}
%}

\toctitle{Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing}
\tocauthor{Eugen Berlin, Marko Borazio, Tobias Grosse-Puppendahl}

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the user's environment to boost activity recognition. This work presents a sensor prototype consisting of an accelerometer and a capacitive proximity sensor that senses the user's activities based on the combined sensor values. We show that our proposed approach of combining both modalities significantly improves the recognition rate for detecting activities of daily living.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living. Therefore, we obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

@@

Capacitive proximity sensing faces the great advantage of being robust against changing lighting conditions and occlusion. Moreover, sensing electrodes can be integrated invisibly into the environment. On the other hand, the exact distance to objects can only be approximated since the object's surface, its conductivity and grounding has influence on the measurement result. A single sensor will thus deliver data that has a certain degree of ambiguity. Due to the nature of capacitive proximity sensors, they can be prone to errors in environments with strong and rapidly changing electric fields. This, however, is usually not an issue when considering activities of daily living.

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{Wimmer2007} who presented a toolkit for capacitive proximity sensing. In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

\begin{figure}[h]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{When the participants entered the apartment, the wristband approached the door knob twice, at the time of opening and closing the door. This fact can be observed in the capacitive proximity data (upper plot) at the beginning and at the end of the activity, whereas the acceleration has no characteristic information (bottom plot).}
	\label{fig:opendoor}
\end{figure}

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

@@

We would like to thank our evaluation participants from TU Darmstadt and Fraunhofer IGD.
